---
title: "Create dataset for neuro project"
author: "MWS"
date: "2023-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(data.table)
source(here("R/functions.r"))
```

# 1. Load data
```{r}
# UKB showcase data
df <- fread(here("data-raw/your_UKB_data.csv"))

# dataset with gcgr_variants
gcgr <- fread(here("data/your_gcgr_data.csv"))

# Withdrawals (updated 14/11 2023)
withdrawals <- fread(here("data-raw/withdraw_data.txt"))


# Neurodegenerative diseases were downloaded from DNA nexus
neuro <- fread(here("data-raw/your_neuro_data.csv"))
setnames(neuro, "Participant ID", "eid")
```

Select variant columns from the gcgr dataset to include in Sashas data
```{r}
setnames(gcgr, "trunc", "frameshift") # we decided to call it framehshift mutations instead of truncated variants
cols_to_keep <- c("eid", "variant_type", "cAMP_LoF", "frameshift", "G40S")
gcgr2 <- gcgr[, ..cols_to_keep]
```

Update col names
```{r}
setnames(df, "body_mass_index_bmi_f21001_0_0", "BMI")
setnames(df, "age_when_attended_assessment_centre_f21003_0_0", "age_enrolment")
setnames(df, "sex_f31_0_0", "sex")
setnames(df, "glycated_haemoglobin_hba1c_f30750_0_0", "hba1c")
```


# 2. Define T2D 
## 2.1 Full period
```{r}
df <- find_ICD10_filter(df, c("E11", "E14"), "T2D")
df <- find_ICD10_filter(df, "E10", "T1D")
count(df, T1D, T2D)
```

```{r}
# Remove diabetes diagnosis if gestational diabetes only
df$gdmonly[df$gestational_diabetes_only_f4041_0_0 == 1 | df$gestational_diabetes_only_f4041_1_0 == 1 | df$gestational_diabetes_only_f4041_2_0 == 1 | df$gestational_diabetes_only_f4041_3_0 == 1] <- T
df$T1D[df$gdmonly == T & df$T2D == T] <- NA
df$T2D[df$gdmonly == T & is.na(df$T1D)] <- NA
df$T1D[df$gdmonly == T & is.na(df$T2D)] <- NA
```

**Diabetes medication codes**
*Metformin*
1140884600	metformin
1140874686  glucophage 500 mg tablet 
1141189090	rosiglitazone 1mg / metformin 500mg tablet

*Sulfonylureas*
1140874718  glibenclamide 
1140874744	gliclazide
1140874746  diamicron 80 mg tablet 
1141152590  glimepiride
1141156984  amaryl 1 mg tablet
1140874646	glipizide
1141157284	glipizide product
1140874652  minodiab 2.5 mg tablet
1140874674  tolbutamide
1140874728  euglucon 2.5 mg

*Others*
1140868902	acarbose
1140868908  glucobay 50 mg tablet
1140857508  glucotard 5g/sachet mini-tablet
1140882964  oral hypoglycemic

*Metglitinides*
1141173882	nateglinide
1141173786  starlix 60 mg tablet
1141168660	repaglinide

*Glitazones*
1141171646	pioglitazone
1141171652  actos 15 mg tablet
1141153254  troglitazone
1141177600	rosiglitazone
1141177606  avandia 4mg tablet

*Insulin*
1140883066	insulin product

Filter for oral antihyperglycemic drugs and insulin
```{r echo=TRUE, results='hide'}
# oral antihyperglycemic drugs
df <- find_drug_filter(df, c("1140874718","1140874744","1140874746","1141152590","1141156984","1140874646","1141157284","1140874652","1140874674","1140874728","1140868902","1140868908","1140857508","1141173882","1141173786","1141168660","1141171646","1141171652","1141153254","1141177600","1141177606","1140882964"), "oral_T2D_drug")

# oral antihyperglycemic drugs at baseline
df <- find_drug_filter_bs(df, c("1140874718","1140874744","1140874746","1141152590","1141156984","1140874646","1141157284","1140874652","1140874674","1140874728","1140868902","1140868908","1140857508","1141173882","1141173786","1141168660","1141171646","1141171652","1141153254","1141177600","1141177606","1140882964"), "oral_T2D_drug_bs")

# Metformin treatment
df <- find_drug_filter(df, c("1140884600","1140874686","1141189090"), "metformin")

# Metformin treatment at baseline
df <- find_drug_filter_bs(df, c("1140884600","1140874686","1141189090"), "metformin_bs")

# Insulin treatment
df <- find_drug_filter(df, "1140883066", "insulin_treatment")

# Insulin treatment at baseline
df <- find_drug_filter_bs(df, "1140883066", "insulin_treatment_bs")
```

Defining diabetes (2023 prevalence) = full period (fp)
```{r}
# create
df$T2D_fp <- df$T2D
df$T1D_fp <- df$T1D

# MAke those individuals wihtout diabetes diagnosis as T2D if hba1c above 48:
df$T2D_fp[df$hba1c>48 & (df$T2D_fp == F & df$T1D_fp == F)] <-  T

# Make those individuals without diabetes diagnosis as T2D if oral T2D medication (excluding metformin):
df$T2D_fp[df$oral_T2D_drug == T & (df$T2D_fp == F & df$T1D_fp == F)] <- T

# Make those T1D individuals on oral T2D drug medication T2D:
df$T1D_fp[(df$T2D_fp == T & df$T1D_fp == T) & (df$oral_T2D_drug == T | df$metformin == T)] <- F
df$T2D_fp[df$T1D_fp == T & (df$oral_T2D_drug == T | df$metformin == T)] <- T
df$T1D_fp[df$oral_T2D_drug == T | df$metformin == T] <- F

# Make those diagnosed <20 y age as T1D
df$T2D_fp[(df$age_diabetes_diagnosed_f2976_0_0 < 20 | df$age_diabetes_diagnosed_f2976_1_0 < 20 | df$age_diabetes_diagnosed_f2976_2_0 < 20 | df$age_diabetes_diagnosed_f2976_3_0 < 20 ) & df$T1D == T & df$T2D == T] <- F

# Make those with double diagnosis as T1D if insulin treatment within 1 year of diagnosis and actual insulin treatment:
df$ins1y[df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_0_0 == 1 | df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_1_0 == 1 | df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_2_0 == 1 | df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_3_0 == 1] <- T 
df$ins1y[is.na(df$ins1y)] <- F
df$T2D_fp[df$T1D_fp == T & df$T2D_fp == T & 
              (df$ins1y & df$insulin_treatment)] <- F

# Making those with double diagnosis T1D and T2D as T2D
df$T1D_fp[df$T1D_fp == T & df$T2D_fp == T] <- F

count(df, T1D, T2D) # only ICD10 codes
count(df, T1D_fp, T2D_fp) # after futher definitions
```

## 2.2 Defining baseline T2D
```{r}
# Full-period prevalent T1 and T2 from ICD codes was performed above

# Rename date columns
setnames(df, "date_e14_first_reported_unspecified_diabetes_mellitus_f130714_0_0", "date_e14_rep")
setnames(df, "date_e11_first_reported_noninsulindependent_diabetes_mellitus_f130708_0_0", "date_e11_rep")

# Remove odd dates for t2d diagnosis (https://biobank.ndph.ox.ac.uk/showcase/coding.cgi?id=819)
codings <- c("1900-01-01", "1901-01-01", "1902-02-02", "1903-03-03", "1909-09-09", "2037-07-07")

df[date_e11_rep %in% codings, date_e11_rep := NA]
df[date_e14_rep %in% codings, date_e14_rep := NA]

# Merge the date columns and keep the earliest date of the two if both E11 and E14 are present
df <- df %>% 
   mutate(t2d_date = pmin(date_e11_rep, date_e14_rep))
df[is.na(date_e11_rep), t2d_date := date_e14_rep]
df[is.na(date_e14_rep), t2d_date := date_e11_rep]

# Create column of baseline t2d 
df[, T2D_bs := 0]
df[T2D == 1 & t2d_date < date_of_attending_assessment_centre_f53_0_0, 
   T2D_bs := 1]

count(df, T2D_bs, hba1c>48)

# Make those individuals wihtout diabetes diagnosis as T2D if hba1c (at baseline) above 48:
df$T2D_bs[df$hba1c>48 & (df$T2D_bs == F & df$T1D_bs == F)] <-  1

count(df, T2D_bs)

# Make those individuals without diabetes diagnosis as T2D if oral T2D medication (excluding metformin) at baseline:
df$T2D_bs[df$oral_T2D_drug_bs == T & (df$T2D_bs == F & df$T1D_bs == F)] <- T

# Make those T1D individuals on oral T2D drug medication T2D:
df$T1D_bs[(df$T2D_bs == T & df$T1D_bs == T) & (df$oral_T2D_drug_bs == T | df$metformin_bs == T)] <- F
df$T2D_bs[df$T1D == T & (df$oral_T2D_drug_bs == T | df$metformin_bs == T)] <- T
df$T1D_bs[df$oral_T2D_drug_bs == T | df$metformin_bs == T] <- F

# Make those diagnosed <20 y age as T1D
df$T2D_bs[(df$age_diabetes_diagnosed_f2976_0_0 < 20 | df$age_diabetes_diagnosed_f2976_1_0 < 20 | df$age_diabetes_diagnosed_f2976_2_0 < 20 | df$age_diabetes_diagnosed_f2976_3_0 < 20 ) & df$T1D_bs == T & df$T2D_bs == T] <- F

# Make those with double diagnosis as T1D if insulin treatment within 1 year of diagnosis and actual insulin treatment:
df$ins1y[df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_0_0 == 1 | df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_1_0 == 1 | df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_2_0 == 1 | df$started_insulin_within_one_year_diagnosis_of_diabetes_f2986_3_0 == 1] <- T 
df$ins1y[is.na(df$ins1y)] <- F
df$T2D_bs[df$T1D_bs == T & df$T2D_bs == T & (df$ins1y & df$insulin_treatment_bs)] <- F

# Making those with double diagnosis T1D and T2D as T2D
df$T1D_bs[df$T1D_bs == T & df$T2D_bs == T] <- F

count(df, T2D_bs)
```

# 3. Alcohol
Calculate sum of average weekly intake of alcohol
```{r echo=TRUE, results='hide'}
# Set prefer not to answer and do not know (-1 and -3) to NA
df$average_weekly_beer_plus_cider_intake_f1588_0_0[df$average_weekly_beer_plus_cider_intake_f1588_0_0<0] <- NA
df$average_weekly_champagne_plus_white_wine_intake_f1578_0_0[df$average_weekly_champagne_plus_white_wine_intake_f1578_0_0<0] <- NA
df$average_weekly_fortified_wine_intake_f1608_0_0[df$average_weekly_fortified_wine_intake_f1608_0_0<0] <- NA
df$average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0[df$average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0<0] <- NA
df$average_weekly_red_wine_intake_f1568_0_0[df$average_weekly_red_wine_intake_f1568_0_0<0] <- NA
df$average_weekly_spirits_intake_f1598_0_0[df$average_weekly_spirits_intake_f1598_0_0<0] <- NA

# units of alcohol in beverages: https://www.nhs.uk/live-well/alcohol-advice/calculating-alcohol-units/

# weekly units of alcohol - redwine (f1568), 1 glass = 2.1 units
df$average_weekly_red_wine_intake_f1568_0_0 <- df$average_weekly_red_wine_intake_f1568_0_0 * 2.1
# weekly units of alcohol - champagne etc (f1578), 1 glass = 2.1 units
df$average_weekly_champagne_plus_white_wine_intake_f1578_0_0 <- df$average_weekly_champagne_plus_white_wine_intake_f1578_0_0 * 2.1
# weekly units of alcohol - beer and cider (f1588), 1 pint = 2 units
df$average_weekly_beer_plus_cider_intake_f1588_0_0 <- df$average_weekly_beer_plus_cider_intake_f1588_0_0 * 2
# weekly units of alcohol - spirits (f1598), 1 measure = 1 unit (no calculation needed)
# weekly units of alcohol - fortified wine (f1608), 1 measure = 1 unit (no calculation needed)
# weekly units of other alcoholic drinks incl alcopops, 1 glass = 1.5 units
df$average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0 <- df$average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0 * 1.5

#Calculate the sum of weekly units alcoholic intake
df <- df %>%
    rowwise() %>%
    mutate(weekly_alcohol = sum(average_weekly_beer_plus_cider_intake_f1588_0_0,
    average_weekly_champagne_plus_white_wine_intake_f1578_0_0,
    average_weekly_fortified_wine_intake_f1608_0_0,
    average_weekly_intake_of_other_alcoholic_drinks_f5364_0_0,
    average_weekly_red_wine_intake_f1568_0_0,
    average_weekly_spirits_intake_f1598_0_0,
    na.rm=T))

```

# 4. Neurodegenrative diseases
```{r}
setDT(df)
df <- find_ICD10_filter(df, c("A810", "F000", "F001", "F002", "F009", "F010", "F011", "F012", "F013", "F018", "F019", "F020", "F021", 
                              "F022", "F023", "F024", "F028", "F03", "F051", "F106", "G300", "G301", "G308", "G309", "G310", "G311", "G318", "I673"),
                        "demens_diseases")

df <- find_ICD10_filter(df, c("F00", "F01", "F02", "F03", "F04", "F05", "F06", "F07", "F09", "G30", "G31", "G32"), "neuro_diseases")

df <- find_ICD10_filter(df, c("F00", "G30"), "alzheimer")
df <- find_ICD10_filter(df, c("F01", "F02", "F03", "F04", "F05", "F06", "F07", "F09", "G31", "G32"), "neuro_non_alzheimer")

```


# 5. Diseases_of_the_pancreas
```{r}
setDT(df)
df <- find_ICD10_filter(df, c("K85", "K86", "K87"), "pancreas_disease")
```


# 6. Age in 2023
```{r}
df$age_2023 <- 2023 - df$year_of_birth_f34_0_0
```

# 7. Rename date for enrolent
```{r}
df$date_enrolment <- df$date_of_attending_assessment_centre_f53_0_0
```

#8. Rename and merge death dates
```{r}
#there are two date collumns. Here they are merged into one collumn, and the earliest is kept in case there is a date in both cols. 
df$date_death <- pmin(df$date_of_death_f40000_0_0, df$date_of_death_f40000_1_0, na.rm = TRUE)

```

#9. Rename lost to follow up
```{r}
df$date_lost_to_followup <- df$date_lost_to_followup_f191_0_0
```


# 10. Filter for ethnicity and sex chromosome aneuploudy
```{r}
# Exclude non-white participants 
count(df, genetic_ethnic_grouping_f22006_0_0)
df <- df[genetic_ethnic_grouping_f22006_0_0 == 1,] # 1 = caucasian

# Exclude individuals with sex chromosome aneuploidy
count(df, sex_chromosome_aneuploidy_f22019_0_0)
df <- df[is.na(sex_chromosome_aneuploidy_f22019_0_0),]
```


# 11. clean up df and merge with gcgr variant data
```{r}
cols_to_keep <- c("eid", "sex", "BMI", "age_enrolment", "age_2023", "date_death", "date_lost_to_followup", "T1D_fp", "T2D_fp", "T2D_bs", "weekly_alcohol", "demens_diseases", "neuro_diseases", "pancreas_disease","alzheimer", "neuro_non_alzheimer", "date_e11_rep", "date_e14_rep", "date_enrolment")
df2 <- df[, ..cols_to_keep]

df_merged <- merge(df2, gcgr2, by = "eid", all.x = TRUE)
df_merged <- merge(df_merged, neuro, by = "eid", all.x = TRUE)

head(df_merged)
```

Export df_merged as csv
```{r}
fwrite(df_merged, file = here("data/your_data_filtered.csv"))
```




